Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-03-29 analysis/paper/paper.Rmd
Updated manuscript

---
title: "The Quartzite Conundrum: Non-destructive analysis of Bronze Age quartzite points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Sweden 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
    \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl"
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. This leads to a severe lack of methodological development in the characterization of these same materials in an archaeological context. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). The data was evaluated with exploratory analysis using Principal Component Analysis. The results demonstrate that detectable differences exist within the dataset for each instrumentation. A number of features identified, such as the presence of graphite, can likely be related to diagenetic/paragenetic differences in the material. Not only should these results aid in future surveys of geological sources in the region, but they will also aid in the construction of predictive models for classifying quartzite.  
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

@@

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources. Flint and obsidian have both been central to the study of trade and exchange networks in European prehistory, due in part to the well-surveyed geological sources for these materials. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al. [-@prieto_etal19 p. 15]. In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region as they represent the most common material groups at archaeological sites. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], in contrast, there has been little development on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the current landscape. Existing theories on lithic material sourcing within the North Swedish region tend to describe people employing strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. Spectroscopic instrumentation features fast measurements that can be collected directly on the surface of an object. With newer mobile instrumentations readily available, this makes it an attractive option for field-based screening purposes. In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other complementary instrumentations that may provide further insight into the structural properties of a material. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

@@

In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig. \@ref(fig:nir-spectra-summary)). The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden. The sites are distributed along the Ångermanälven river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south. A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. No distinction is made between arrow- and spearheads.

A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, @mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].

@@

```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine. Made with ©SGU, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
knitr::include_graphics(here::here("analysis", "figures", "site_distribution_soils_annotated.png"), error = FALSE)
```

@@

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24]. As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

```{r sample-hue-pictures, fig.cap= "Not only is the material heterogeneous in terms of colour, but the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation."}
knitr::include_graphics(here::here("analysis", "figures", "figure_2.png"), error = FALSE)
```

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/]. The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens. The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape. As the study is non-destructive sampling was performed on the surface of a material that is not uniform. The material ranges from fairly small and flat point fragments, to large and rough preforms. Therefore the integration could at times be difficult, which in turn could lead to scattering effects and saturation.

Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. When sampling materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

@@

(ref:artefact) Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts. Note that the term ""point"" does not differentiate between spearheads or arrowheads.

@@

## Instrumentaion and pre-processing {-}

@@

### Near Infrared {-}
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77 p.508].

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå^[www.slu.se/en/departments/forest-biomaterials-technology/].

After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using modified polynomial fitting [@lieber_mahadevan-jansen03] from the ChemoSpec package [@hanson16]. The spectra were then mean-centred <!--and normalized using Standard Normal Variate [@shaver01 pp.284-285] --> before PCA.

@@

### X-Ray Fluourescence {-}
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Moseley's law [@moseley1913] recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms. This is what enables the spectral features in XRF spectrometry to be related to certain elements. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12 pp. 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was *mining* mode [@knight_etal21 p.2]. The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light). Due to time constraints, and the number of artefacts included in the project, it was only possible to sample each object once, apart from some exceptions.

It is also important to address the *closed sum* effect of compositional data [@pawlowsky-glahn_egozcue06]. As the variables are given as proportions of a constant sum (e.g. percentage, parts per million) this means that if one element was to change in value, others would follow [@grunsky_caritat20]. This is managed by transforming the data using log-ratio transformation [@aitchison99], and currently there are three main methods for achieving this; additive log-ratio (alr), centered log-ratio (clr) or isometric log-ratio. They all have their own nuances and as such the choice of method is dependent on the type of analysis and material [@pawlowsky-glahn_etal15]. In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.

@@

The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20]. Multivariate approaches, such as PCA,  has proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predicitive models.

@@

#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
points.d.filt <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
                           "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56")) %>% 
  filter(hue == "Dark")

#pass to vector so that length() can be used in text to quote nr. of dark samples
spec.d.total <- 
  points.d.tot$sample_id

#pass to vector so that length() can be used in text to quote nr. of dark samples
spec.d.filtered <- 
  points.d.filt$sample_id
```

@@

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21 pp. 351-397; @schmid_etal22]. An example of this can be seen in fig. \@ref(fig:nir-savitzky-transform), where the spectra from two different dark objects are compared. The non-processed spectra are fairly indistinct, but the derivative transformation of sample 258 amplifies a number of significant peaks. The derivative spectra of sample 411, however, does not have as clear and distinct peaks and features mainly noise. In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation. After reviewing the dark spectra, out of a total of `r length(spec.d.total)`, `r length(spec.d.filtered)` were kept in the dark group.

@@

```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca.R")), encoding = "UTF-8", echo = FALSE}
quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
nir.pc1var <- nir.pc1var
nir.pc2var <- nir.pc2var
nir.pc3var <- nir.pc3var
```

@@

# CRediT authorship contribution statement {-}
Mattias Sjölander: Conceptualization, Data curation, Methodology, Software, Validation, Formal analysis, Resources, Writing - original draft, Visualization, Project administration.

Johan Linderholm: Conceptualization, Methodology

>>>>

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Sweden 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl"
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of the same these materials. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin and potentially the metamorphic grade. These features indicate the presence of graphite, muscovite, and biotite. Apart from reaffirming the potential of field-based screening instrumentation, these results should aid in future surveys of geological sources in the region. There also seems to be a good ground for the construction of a predictive model which that could classify the quartzite based on its chemical characteristic. Such a model could prove useful in spatial analysis and testing of models of raw material management.
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

@@

Petrographic characterization and provenance studies are instrumental for inferences of trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent has been highlighted by Prieto et al. [-@prieto_etal19: 15]. In Northern Sweden both quartz and quartzite dominate the lithic assemblages and have been used in studies of settlements, exchange and mobility [@baudou78]. While quartz and quartzite technology have been the target of several studies over the years, there has been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. M While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden haves been put forth [@halen94: 115-122; @lundberg97: 161-171]. On the other hand there haves been little development on the topic of quartz and quartzite [@holm91: 23-24]. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine deposits and the degree to which geological materials have been mixed and transported necessitates considerable effort to reconstruct the events that have formed the current landscape. Existing theories on lithic material sourcing within the North Swedish region tend to describe strategies regarding direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06: 249-250]. Due to this ambiguous nature of availability of raw materials, provenance studies are complex.

Prieto et al. [-@prieto_etal19: 15; -@prieto_etal20: 32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature. Because of this , and different researchers have thus emphasized different different features as being characteristics of the material [@howard05]. Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05: 708] also emphasises the need for field-based methodologies. Spectroscopic surface measurements are fast, and can be adapted for use in the field. In the case of archaeological research there are situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.

X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03]), due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF, as described by Norris [-@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

@@

In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig. \@ref(fig:distribution-map)). The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.

@@

The current study includes spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden (fig. \@ref(fig:distribution-map)).The sites are distributed along the Ångerman river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south.  The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. No distinction is made between arrow- and spearheads.

A number of these points have, according to prior classifications, been worked using bifacial technology. The geographically closest chronology for bifacial points can be found in southeast Norway [@mjaerum12], but so far few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89], ). These attempts suggests a period of use between ca. 3 800 - 2 000 BP [@kaeck09: 60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].

@@

```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
knitr::include_graphics(here::here("analysis", "figures", "site_distribution_soils_annotated.png"), error = FALSE)
```

@@

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91: 24]. As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

```{r artefacts, fig.cap= "The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 2-76 mm. "}
knitr::include_graphics(here::here("analysis", "figures", "figure_2.png"), error = FALSE)
```

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/). The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens. The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape (fig. \@ref(fig:artefacts)-\@ref(fig:artefact-size)) [see Appendix A](#appendix-a). The study is non-destructive sampling was performed on the surface of a material that is not uniform. The material ranges from fairly small and flat point fragments, to large and rough preforms. Therefore, the signal integration could at times be difficult, which in turn could lead to scattering effects and saturation.

Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. Sampling materials that featured high heterogeneity the sample point was focused on the surrounding matrix, away from deviating inclusions. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

@@

(ref:artefact) Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts. 8 objects have no recorded thickness, which ranges between 2-76 mm. Note that the term *point* does not differentiate between spearheads or arrowheads.

@@

## Instrumentation and data pre-processing {-}

@@

### Near Infrared {-}
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near infrared region. The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify the bond. Overtones occur at frequencies that are a multiple of the infrared vibration. Most of them end up in the near infrared region [@osborne_etal93].

Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508].

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^. After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using Lieber and Mahadevan-Jansens' [-@lieber_mahadevan-jansen03] algorithm for polynomial fitting, with the ChemoSpec package [@hanson16]. The spectra were then mean-centred before PCA.

@@

### X-Ray Fluourescence {-}
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Spectral features in XRF spectrometry can be related to certain elements according to Moseley's law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24]. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was *mining* mode [@knight_etal21: 2]. The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light). Due to time constraints, and the number of artefacts included in the project, it was only possible to sample each object once, apart from some exceptions.

As the variables in XRF analysis is given as proportions of a constant sum (e.g. percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06]. This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20]. This has implications for classical statistical methods, and is managed by transforming the data using log-ratio transformation [@aitchison99]. There are currently three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15]. In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.

@@

The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20]. Multivariate approaches, such as PCA,  were proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predictive models.

@@

#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
points.d.filt <-
  points.d.tot %>% 
  filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
                           "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56"))

#pass to vector so that length() can be used in text to quote total nr. of samples
spec.d.total <- 
  points.d.tot$sample_id

#pass to vector so that length() can be used in text to quote nr. of dark samples after SG analysis
spec.d.filtered <- 
  points.d.filt$sample_id
```

@@

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17: 78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21: 351-397; @schmid_etal22]. An example of this can be seen in fig. \@ref(fig:nir-savitzky-transform), where the spectra from two different dark objects are compared. The non-processed spectra are fairly indistinct, but the derivative transformation of sample 258 amplifies a number of significant peaks. The derivative spectra of sample 411, however, does not have as clear and distinct peaks and features mainly noise. In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation. After reviewing the dark spectra, out of a total of `r length(spec.d.total)`, `r length(spec.d.filtered)` were kept in the dark group.

@@

```{r nir-pca, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca.R")), encoding = "UTF-8", echo = FALSE}
quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
nir.pc1var <- nir.pc1var
nir.pc2var <- nir.pc2var
nir.pc3var <- nir.pc3var
```

@@

# CRediT authorship contribution statement {-}
Mattias Sjölander: Conceptualization, Data curation, Formal analysis, Methodology, Project administration, Resources, Software, Visualization, Validation, Writing - original draft.

Johan Linderholm: Conceptualization, Methodology, Supervision
